---
title: "homework1-stat154"
output:
  html_document:
    df_print: paged
---

Problem 1 (10 pts)

Create the following matrices in R (and display them).

```{r}
X<-matrix(c(2,-1,3,4),ncol=2,nrow=2)
print(X)
```

```{r}
Y<-matrix(c(2,1,0,-2,1,3),ncol=3,nrow=2)
print(Y)
```

```{r}
Z<-matrix(c(1,-1,0,1,1,2),ncol=2,nrow=3)
print(Z)
```

```{r}
W<-matrix(c(1,8,0,3),ncol=2,nrow=2)
print(W)
```

```{r}
I<-diag(x=1,nrow=2,ncol=2)
print(I)
```

Problem 2 (10 pts)

Use the matrices created in Problem 1 to perform each of the following operations in R. If the indicated operation cannot be performed, explain why.

a. X+Y 

  since X and Y are non-conformable arrays, so they can not do add operation.

b. X+W

```{r}
X+W
```

c. X−I

```{r}
X-I
```

d. XY

```{r}
X%*%Y
```

e. XI

```{r}
X%*%I
```

f. X+(Y+Z) 

  since Y and Z are non-conformable arrays, so they can not do add operation.

g. Y(I+W) Error 

  since I and W are non-conformable arrays, so they can not do add operation.


Problem 3 (10 pts)

Determine whether the following statements are True or False. 

a. Every orthogonal matrix is nonsingular.
  True

b. Every nonsingular matrix is orthogonal.
  True
c. Every matrix of full rank is square.
  False

d. Every square matrix is of full rank.
  False

e. Every nonsingular matrix is of full rank.
  True
Problem 4 (10 pts)

Let X, Y, and Z be conformable. Using the properties of transposes, prove that:
(XYZ)T = ZTYTXT

prove

answer see the picture.

Problem 5 (10 pts)

Consider the eigenvalue decomposition of a symmetric matrix A. Prove that two eigenvectors vi and vj associated with two distinct eigenvalues λi and λj of A are mutually orthogonal; that is, viTvj = 0

answer see the picture.

6.1 Function inner_product (10 pts)

```{r}
inner_product<-function(v,u) {
  res<-sum(v*u);
  return(res);
}
v<-c(1,3,5)
u<-c(1,2,3)
inner_product(v,u)
```

6.2 Function projection() (10 pts)

```{r}
projection<-function(v,u) {
  res<-inner_product(u,v)/inner_product(u,u)*u;
  return(res);
}
projection(v,u)
```

Problem 7 (10 pts)

```{r}
x<-c(1,2,3)
y<-c(3,0,2)
z<-c(3,1,1)
u1<-x
u2<-y
u3<-z
e1<-projection(u1,u1)
e2<-projection(u2,u1)
e3<-projection(u3,u1)
```

Problem 8 (10 pts)

```{r}
lp_norm<-function(x,p=1) {
  if (p=="max") {
    res <-max(abs(x))
    return(res)
  }
  res<-sum(abs(x)^p)^(1/p);
  return(res);
}
```

Problem 9 (10 pts)

Use your function lp_norm() with the following vectors and values for p: 
a. zero <- rep(0, 10)andp = 1
b. ones <- rep(1, 5)andp = 2
c. u <- rep(0.4472136, 5) and p = 2
d. u <- 1:500andp = 100
e. u <- 1:500 and p = "max"

```{r}
zero <- rep(0, 10)
p<-1
lp_norm(zero,p)
```

```{r}
ones <- rep(1, 5)
lp_norm(ones,p=2)
```

```{r}
u <- rep(0.4472136, 5)
lp_norm(u,p=2)
```

```{r}
u <- 1:500
lp_norm(u,p=100)
```

```{r}
lp_norm(u,p="max")
```

Problem 10 (10 pts)

Consider the eigendecomposition of a square matrix A.
a. Prove that the matrix bA, where b is an arbitrary scalar, has bλ as an eigenvalue, with v as the associated eigenvector.
b. Prove that the matrix A+cI, where c is an arbitrary scalar, has (λ+c) as an eigenvalue, with v as the associated eigenvector.

answer see the picture.

Problem 11 (20 pts)

For this problem, use the data set state.x77 that comes in R.
a. Select the first five columns of state.x77 and convert them as a matrix; this will be the data matrix X. Let n be the number of rows of X, and p the number of columns of X

```{r}
X<-as.matrix(state.x77[,1:5])
n<-nrow(X)
p<-ncol(X)
```

b. Create a diagonal matrix D = 1/n I where I is the n × n identity matrix. Display the output of sum(diag(D)).
```{r}
I<-diag(x=1,nrow=n,ncol=n)
D<-(1/n)*I
sum(diag(D))
```
c. Compute the vector of column means g = XTD1 where 1 is a vector of 1’s of length n.
Display (i.e. print) g.
```{r}
L<-rep(1,nrow(X))
g<-t(X)%*%D%*%L
print(g)
```
d. Calculate the mean-centered matrix Xc = X−1gT. Display the output of
colMeans(Xc).
```{r}
Xc<-X-(L%*%t(g))
colMeans(Xc)
```

e. Compute the (population) variance-covariance matrix V = XTDX − ggT. Display the
output of V.
```{r}
V<-t(X)%*%D%*%X-g%*%t(g)
V
```
f. Let D1/S be a p × p diagonal matrix with elements on the diagonal equal to 1/Sj , where Sj is the standard deviation for the j-th variable. Display only the elements in the diagonal of D1/S
```{r}
cal_sj<-function(p,X) {
  S<-vector(,p)
  for (i in 1:p) {
     S[i]<-sd(X[,i])
  }
  return(S)
}
Sj<-cal_sj(p,X) 
Ds<-diag(x=1/Sj,nrow=p,ncol=p)
print(1/Sj)
```
g. Compute the matrix of standardized data Z = XcD1/S Display the output of colMeans(Z) and apply(Z, 2, var)
```{r}
Z<-Xc%*%Ds
print(colMeans(Z))
apply(Z,2,var)
```

h. Compute the (population) correlation matrix R = D1/SVD1/S. Display the matrix R
```{r}
R<-Ds%*%V%*%Ds
print(R)
```
i. Confirm that R can also be obtained as R = ZTDZ
```{r}
R<-t(Z)%*%D%*%Z
print(R)
```

Comments and Reflections

Reflect on what was hard/easy, helpful tutorials you read, etc. Here are some questions that you can start with to provide comments on your HW.

• What things were hard, even though you saw them in class?
eigendecomposition and singular decomposition.

• What was easy(-ish) even though we haven’t done it in class?
create matrix.

• Did you need help to complete the assignment? If so, what kind of help? Who helped you?
I have asked TA for questions. 

• How much time did it take to complete this HW?
almost 20 hours

• What was the most time consuming part?
I took a long time to understand the meaning of question.